Grade Retention and Mixed Race

### 2022-08-31 products/paper/main.qmd
Full rough draft of biracial fit section

To better understand where biracial individuals might fit, it is necessary to understand what is producing a contemporary system of racial inequality overall. Here, I distinguish three mechanisms that may help us understand the biracial position within this racial hierarchy. First, I consider arguments for the racial hierarchy that focus on the effects of racial discrimination at both an individual and systemic level. Second, I consider arguments regarding different distributions of resources across racial groups that produce different average arguments. Thirs, I consider arguments regarding how the nature of hybridity itself may affect outcomes.

@@

Possible restructuring of this whole section:

* To understand where biracials fit, it is helpful to think about what mechanisms drive the racial hierarchy more broadly.
  * First, racial inequality in outcomes may be produced by the cumulative sum of individual acts of discrimination as well as systematic forms of bias and discrimination. Individual discrimination need not be consciously performed and may instead reflect cognitive biases and schema that individuals unknowingly rely upon that produce differential outcomes by race. This discrimination will be more consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers. 
  * Systematic discrimination, on the other hand, reflect biases built into institutional systems that channel rewards to certain racial groups. This can operate through such things as norms, laws, and organizational practices. In our current "colorblind" era, such mechanisms may be formally race neutral, but produce disparate impacts by race. 
  * In either case, the experiences of biracial individuals within such a system of discrimination depend on how they are classified by others. To the extent that biracial individuals are classified as more like one of their monoracial constituent groups. they will receive the same treatment and either reap similar rewards or suffer similar discrimination and penalties.
  * discussion of hypodescent/hyperdescent as governing classification into single-race groups. 
  * Iversen et. al.'s critique of hypodescent/hyperdescent, higlighting:
    * historical specificy of hypodescent and hyperdescent pattern
    * equally consistent with a "succession/recession" perspective
    * regimes governing certain groups may differ due to historical circumstances.
  * How does this questions of classification address the case where biracial individuals might be in-between? A couple of possibilities.
    * Racial ambiguity could lead to more inconsistent classification into single race categories across observers submitting biracials to differing exposure to advantages/distadvantages somewhere between  their two monoracial constituent groups.
    * Greater physical ambiguity may allow biracials to avoid worst aspects of discrimination because they don't match the "ideal type" for most discriminated group. We already know that this process exists among monoracial blacks in terms of skin tone, so this is not unique to multiracials.
    * It is also possible that biracials are "correctly" classified by observers into a "middleman" minority position. There is evidence of this kind of practice in Latin America, and it may become more common in the US (Bonilla-Silva). It is unlikely that observers are capable of differentiating fine divisions of ancestry but rather can put people into a nebulous middleman position.
* Some racial inequality may also be explained by differential distributions of resources across racial groups. On average, non-whites come from a lower SES background than whites and this may indirectly lead to observed racial differences. Although much of the focus of the "race vs. class" debate is on SES characteristics, less concretely defined social and cultural capital differences across race may exacerbate racial inequality.
  * The role of these resource differences in biracial's position depends on what resources the typical parent in an interracial union brings with them. This is a complex and not-well understood process even for easily measurable characteristics like education. ...
  * the black box of symbolic resources.
* Aside from these major mechanisms for producing racial inequality, the experience of hybridity itself may affect the outcomes of biracial individuals relative to their monoracial peers.
  * Marginal man theory
  * Enlarged toolkits

The most direct pathway for such placement is via how biracial individuals are classified by those in positions of power and control, such as teachers, administrators, and employers. To the extent that biracial individuals are classified as more like one of their monoracial constituent groups. they will receive the same treatment and either reap similar rewards or suffer similar discrimination and penalties.

Two well known norms have long governed such patterns of classification. The pattern of *hypodescent* dictates that individuals with mixed racial ancestry will be classified as members of the lower status group while the pattern of *hyperdescent* dictates that such individuals will be classified as members of the higher status group (==CITATIONS==). Hypodescent, in the form of the "one drop rule" has long characterized the experiences of mixed part-Black individuals who are treated as exclusively Black (==CITE F JAMES DAVIS==). The pattern of hyperdescent blends mixed race individuals into the often unmarked higher status group and is thus more difficult to see in practice. Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming to be White (==CITATIONS==). ==NEED TO MIX IN SOME OF THE COMMENTS IN ANN AND ALIYAS RECENT PAPER==

Norms of hyperdescent and hypodescent may account for cases where the biracial group has outcomes similar to one of the monoracial constituent groups. However, two issues may diminish the relevance of these norms for understanding the placement of contemporary biracial populations. First, for individuals with Asian and Latino ancestry, no historical norm governs the placement of individuals with mixed race. Second, the historical norms for those of part-Black and part-Indigenous ancestry may also be breaking down (CITATION). MORE TO DO HERE In both of those cases, it is not entirely clear if such norms exist outside of cases in which the other component in the mixing is Whiteness. Thus, while these norms may give us some historical expectations for certain groups, they are of limited predictive value more generally.

Patterns of hyperdescent and hypodescent expects biracial individuals to have outcomes more similar to one monoracial group. Another possibility is that biracial individuals will have outcomes that are somewhere between their two constituent monoracial groups. Why might such an in-between status exist? There are several possibilities.

Racial ambiguity in physical appearance may help to produce an in-between status. This racial ambiguity may protect biracial respondents from the worst stigmatization and discrimination in social settings. However, by the same coin, it may allow them to benefit less from membership in advantaged groups. The mechanisms for this process may be two-fold. Racial ambiguity may "soften" any particalar experience of discrimination because the person in question does not conform to an ideal type and therefore may be less likely to trigger a sense of racial threat or antipathy. Additionally, racial ambiguity may contribute to more probabilistic and uncertain assessments of racial identification by various social actors that contribute to a lower sum total of disadvantage.

Notably, both of these mechanisms could co-exist with norms of hypo/hyperdescent. We know, for example, that lighter skin tone among monoracial Black individuals is associated with better outcomes (==CITATION==). So, while the one-drop rule provides a consistent classification system, it does not create uniform experiences for all Blacks and skin tone may help to "soften" the experience of discrimination. (==CITE MONK HERE==)

Aside from ambiguity in physical appearance, a variety of non-material symbolic resources (e.g. social and cultural capital) may also play a role in producing an in-between status. We know that racial disparities persist across a variety of outcomes even after controlling for a wide variety of socioeconomic variables (==CITATIONS==). The residual effect of race here is a "black box" that we often think of as compromising some combination of (a) discrimination and (b) other forms of social and cultural capital not well captured in our data that are differently distributed by race. Biracial individuals may inherit this black box of unobserved resources in weights related to their parent's races, thus helping to produce results that are in-between the two constituent groups. (==I SHOULD GIVE SOME EXAMPLES AND CITATIONS OF WHAT I AM TALKING ABOUT HERE==)

The experience of hybridity itself may also positively and/or negative affect the outcomes for biracial individuals. On the positive side, hybridity may enlarge the cultural "toolkit" (==CITATION==) of biracial individuals. This enlarged toolkit may allow such individuals to more easily navigate a variety of social situations (e.g. code switch). On the negative side, Stonequist's (==CITATION==) "marginal man" hypothesis suggests that the dissonance of an identity that crosses strong and salient racial boundaries will be experienced by biracial individuals as not fitting in anywhere. This liminality may result in negative mental health experiences for biracial individuals that can then affect other types of outcomes as well.

==NO FORMAL HYPOTHESIS==

## Selection into Intermarriage

The placement of biracial individuals may also be affected by mechanisms that drive selection into interracial unions in the previous generation. Such selection may lead to differences on a variety of parental background characteristics between biracial individuals and their monoracial peers. How this selection will affect a biracial child's outcomes for a given type of pairing are complex because it depends on the strength and direction of that selection on each partner, its consistency across different types of unions, and how strongly the selected characteristic predicts outcomes in the next generation.

Most research on such selection has focused on socioeconomic characteristics. Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage (CITATION). Using education as the socioeconomic parameter, prior research has shown strong evidence of this pattern in the case of Black/White marriages in the US (CITATION). Evidence of this pattern for other forms of intermarriage is less clear (CITATION).

Alternatively,

1.  its complicated
2.  Rather than sort out a priori predictions, its important to control for most likely selected characteristics. In the case I explore here socioeconomic status and traits measuring assimilation such as foreign-born status and english language proficiency.
3.  This is not a "race vs. class" argument that diminishes the importance of race. Rather it delineates where generationally and by what mechanisms race might matter for biracial individuals.

------------------------------------------------------------------------

The aforementioned mechanisms all relate to

In addition to the aforementioned racial mechanisms that may drive the placement of biracial individuals, researchers must also pay careful attention to how socioeconomic background may affect the placement of biracial individuals within a racial hierarchy. Accounting for socioeconomic difference often reduces the magnitude of observed racial inequality on a given outcome, suggesting that some of the racial inequality we observe operates through class differences (CITATION). This issue is particularly important for biracial individuals because of potential class selection into interracial unions. The likelihood of being an interracial union varies by class background for at least some groups (CITATION), although it varies in complex ways and we know more about interracial marriages than other types of unions that may produce children. For example, black partners tend to be positively selected with regard to education into Black/White marriages while white partners are negatively selected (CITATION). According to status exchange theory (CITATIONS), this pattern may be common to all types of interracial marriages within an unequal status hierarchy, although the evidence of status exchange for other forms of marriages is weaker (CITATION). How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena for a given pairing, the gender balance in that pairing, and how the gender of parents affects outcomes.

Regardless of the underlying complexity, delineating the role of this selection process in the prior generation as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals into a racialized system. Although this approach is sometimes characterized as pitting race vs class and minimizing the importance of race, that is not case here. Instead, what this delineation allows us to do is understanding how race affects the outcomes of biracial individuals in a multigenerational process and allow us to better pinpoint the generational locus of certain mechanisms.

>>>>

To better understand where biracial individuals might fit, it is necessary to understand what is producing a contemporary system of racial inequality overall. Here, I distinguish three mechanisms that may help us understand the biracial position within this racial hierarchy. First, I consider arguments for the racial hierarchy that focus on the effects of racial discrimination at both an individual and systemic level. Second, I consider arguments regarding different distributions of resources across racial groups that produce different average outcomes. Third, I consider arguments regarding how the nature of hybridity itself may affect outcomes.

## Discrimination and Classification

@@

## Differential Resources

Racial inequality may also be at least partly explained by different distributions of material, cultural, and social resources across racial groups. On average, non-whites come from a lower socioeconomic background (e.g. parental income, wealth, and education) than whites and such socioeconomic background is consequential for outcomes. Other less materially defined resources also vary across groups such as cultural styles, social networks, and neighborhood quality. When such characteristics are held constant or "controlled", the remaining differences between racial groups tend to be smaller, although such differences rarely disappear. These differences in background resources reflect the accumulation of historical processes of discrimination. Thus, observed racial inequality reflects both contemporary processes of discrimination and bias and the intergenerational accumulation of disadvantage.

How do these differential distributions of resources affect biracials? A naive argument would be that because biracial have parents from two different racial groups, they will on average inherit parental resources roughly halfway in-between their two constituent groups (CITE CHEW ET. AL.). Prior research looking at biracial children in the Census data has found some evidence to support this argument (CITE CHEW ET. AL.). If such were the case, then the resources of biracial individuals would predict outcomes in-between the two constituent groups, barring other mechanisms.

However, the actual resource position of biracial individuals is likely far more complicated because individuals are not randomly selected into interracial unions. How this selection will affect a biracial child's outcomes for a given type of pairing are complex because it depends on the strength and direction of that selection on each partner, its consistency across different types of unions, and how strongly the selected characteristic predicts outcomes in the next generation.

Status exchange theory predicts that given an unequal racial hierarchy, individuals from a lower status racial group will be positively selected on socioeconomic characteristics into intermarriage with a higher status group, while individuals from a higher status racial group will be similarly negatively selected into the same marriage (CITATION). Using education as the socioeconomic parameter, prior research has shown strong evidence of this pattern in the case of Black/White marriages in the US (CITATION). Evidence of this pattern for other forms of intermarriage is less clear (CITATION), as is research on other forms of interracial unions such as cohabitations (CITATIONS). In any case, because status exchange positively selects one parent and negatively selects the other partner, the two effects may somewhat cancel each other out in terms of the overall distribution of resources.

Regardless of the underlying complexity of this selection process, delineating its role as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals within the racial hierarchy. Although this approach is sometimes characterized as pitting race versus class and minimizing the importance of race, that is not case here. Instead, what this delineation allows us to do is understand how race affects the outcomes of biracial individuals in a multigenerational process to better pinpoint the generational locus of specific mechanisms.

## Hybridity

The experience of hybridity itself may also positively and/or negatively affect outcomes for biracial individuals. The "marginal man" theory argues that biracial individuals will find it difficult to fit in because of the dissonance of an identity that crosses strong and salient racial boundaries (CITE PARK AND STONEQUIST). This liminal identity conflict will then result in negative mental health experiences that can feed into other negative outcomes as well (CITE CHENG AND LIVELY?).

Marginal man theory, developed in the first half of the Twentieth century, saw cultures as monolithic and static and thus to be trapped in-between could be a difficult experience. Contemporary notions of culture emphasis its "toolkit" nature, asserting a more flexible and fluid notion of culture (CITATION). In this case, biracial individuals may benefit from an enlarged cultural toolkit that allows them to navigate a variety of diverse social contexts (i.e. code switch).

The complexity of the underlying mechanisms described above prevents simple an straightforward hypothesis regarding the position of biracial individuals. In part, this is difficult because similar mechanisms can produce the same outcome. The weight of different mechanisms may also vary by groups as these groups may governed by different regimes of classification (CITATION). However, comparing outcomes across biracial groups may help us better understand the likely mechanisms that drive biracial classification and to disregard others as implausible. Accounting for the effect of parental resources on outcomes to the best of our ability is also key to reaching any conclusions about the placement of biracials within the racial hierarchy.
